---
title: "A report on Lake Ontario's microbes"
author: "Daphne Garcia"
date: "2025-02-19"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#Prepare the R environment

```{r packages}
#Load libraries/packages
library(tidyverse)

```


#Load in the Lake Ontario Data

```{r Load Data}
# load in Lake Ontario microbial community data
sample_and_taxon <-
  read_csv("data/sample_and_taxon.csv")

#inspect the data
glimpse(sample_and_taxon)
```


#Plot Lake Ontario microbial abundance vs temperature
```{r cells_vs_temp}
ggplot(data = sample_and_taxon,
       aes(x=temperature, y=cells_per_ml)) +
  geom_point(aes(color=env_group)) +
  labs(title="Lake Ontario microbial abundance vs temperature",
       x = "Temperature (C)", y= "Cell Density (cells/mL)") +
  theme_classic() +
  theme(legend.position = "bottom", 
        legend.title = element_blank()) +
  scale_color_manual(values=c("lightblue",
                              "gold",
                              "orchid")) +
  geom_smooth(method = "lm", color="black")


```
The above plot shows that:

- there's positive relationship between temperature and cell abundances

- for example: deep samples are the coldest and have the fewest files


```{r data-summary, echo=FALSE}
#How many samples do I have?
n_samples <- 
  sample_and_taxon %>%
  #only pick sample id column
  select(sample_id) %>%
  #number of distinct names in column
  n_distinct()

#calculate the min and max temps
#min
minTemp <- sample_and_taxon %>% 
  summarize(round(min(temperature), digits = 2)) %>%
  pull()
#max
maxTemp <- sample_and_taxon %>% 
  summarize(round(max(temperature), digits = 2)) %>%
  pull()


```


The total number of samples is `r n_samples`. For this set of samples, temperatures range from min of `r minTemp` to max of `r maxTemp`

